Supplementary Material for \\Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-08-03 paper/ifo-appendix.tex

\section{Detecting a sinusoidal signal in Gaussian noise}
\label{app:sinusoid_likelihood}
%\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}
In this appendix, we demonstrate the modulus of the Fourier transform is an appropriate detection statistic when searching for a sinusoidal signal in Gaussian noise. 
We describe the data as
\begin{equation}
x(t) = s(t) + n(t)\,, 
\label{eqn:GNdata}
\end{equation}
where $s(t)$ and $n(t)$ are the signal and noise, respectively and  
\begin{equation}
s(t) = A \cos\left[{\omega t + \phi}\right]\,,
\label{eqn:GNmodel}
\end{equation}
where $A$, $\omega$, and $\phi$ are the amplitude, angular frequency and phase of the signal, respectively.

The likelihood is proportional to the probability of observing the data given an assumed model of the signal~\cite{Jaynes:2003}.
In this example the model is given by Eqn.~\ref{eqn:GNmodel}.
Under the assumption of Gaussian noise, the log-likelihood is 
\begin{equation}
\log \mathcal{L} = -\frac{1}{2}\ip{n}{n}\,, 
\end{equation}
where we define the inner product between $u(t)$ and $v(t)$ as $\ip{u}{v} = \int_0^T \mathrm{d} t \, u(t) v(t) $ and $T$ is the total time of the data. 
We replace $n$ by $x-s$ (Eqn.~\ref{eqn:GNdata}), so that 
\begin{eqnarray}
\log \mathcal{L} &=& -\frac{1}{2}\ip{x-s}{x-s}\,, \\ 
                 &=& -\frac{1}{2}\ip{x}{x} - \frac{A^2}{2} \ip{\cos[\omega t + \phi]}{\cos[\omega t + \phi]} \nonumber\\
                 &&+ A \ip{x}{\cos[\omega t + \phi]}\,,\\ 
                 &=& -\frac{1}{2}\ip{x}{x} - \frac{A^2}{4} + A \ip{x}{\cos[\omega t + \phi]}\,,\label{eqn:logLike}
\end{eqnarray}

@@

%\subsection{Pseudocode}
Here, we present some pseudocode of the implementation used in Section~\ref{sec:viterbi}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. To avoid numerical underflow we take the logarithm and consider them instead as additive weights.

Let $X$ be grid of $j=0,\ldots,N_f$ rows and $i=0,\ldots,N_t$ columns of additive weights for each node. Let Y and Z be grids of the same shape to store the weight of the best path to each node and the row index of the previous node on that path, respectively. Let W be a length $N_t+1$ array to store the final sequence of row indices for the optimal path. We restrict paths to only move one cell up or down at a time (or stay constant). For the edge cases of $j=0,N_f$, we only search over $k \in \{0,1\}$ and $k \in \{-1,0\}$, respectively, to stay inside the grid (this is not shown in the pseudocode below).

>>>>

\section{Detecting a sinusoidal signal in Gaussian noise}
\label{app:sinusoid_likelihood}
%\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}
In this appendix, we demonstrate that the modulus of the Fourier transform is an appropriate detection statistic when searching for a sinusoidal signal in Gaussian noise. 
We describe the data as
\begin{equation}
x(t) = s(t) + n(t)\,, 
\label{eqn:GNdata}
\end{equation}
where $s(t)$ and $n(t)$ are the signal and noise, respectively and  
\begin{equation}
s(t) = A \cos\left[{\omega t + \phi}\right]\,,
\label{eqn:GNmodel}
\end{equation}
where $A$, $\omega$, and $\phi$ are the amplitude, angular frequency and phase of the signal, respectively.

The likelihood is proportional to the probability of observing the data given an assumed model of the signal~\cite{Jaynes:2003}.
In this example the model is given by Eqn.~\ref{eqn:GNmodel}.
Under the assumption of Gaussian noise, the log-likelihood is 
\begin{equation}
\log \mathcal{L} = -\frac{1}{2}\ip{n}{n}\,, 
\end{equation}
where we define the inner product between $u(t)$ and $v(t)$ as $\ip{u}{v} = \int_0^T \mathrm{d} t \, u(t) v(t) $ and $T$ is the total time of the data. 
% We replace $n$ by $x-s$ (Eqn.~\ref{eqn:GNdata}), so that 
Using Eqns.~\ref{eqn:GNdata} and~\ref{eqn:GNmodel}, we find that
\begin{eqnarray}
\log \mathcal{L} &=& -\frac{1}{2}\ip{x-s}{x-s}\,, \\ 
                 &=& -\frac{1}{2}\ip{x}{x} - \frac{A^2}{2} \ip{\cos[\omega t + \phi]}{\cos[\omega t + \phi]} \nonumber\\
                 &&+ A \ip{x}{\cos[\omega t + \phi]}\,,\\ 
                 &=& -\frac{1}{2}\ip{x}{x} - \frac{A^2}{4} + A \ip{x}{\cos[\omega t + \phi]}\,,\label{eqn:logLike}
\end{eqnarray}

@@

%\subsection{Pseudocode}
Here, we present some pseudocode (below) of the implementation used in Section~\ref{sec:viterbi}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. To avoid numerical underflow we take the logarithm and consider them instead as additive weights.

Let $X$ be grid of $j=0,\ldots,N_f$ rows and $i=0,\ldots,N_t$ columns of additive weights for each node. Let Y and Z be grids of the same shape to store the weight of the best path to each node and the row index of the previous node on that path, respectively. Let W be a length $N_t+1$ array to store the final sequence of row indices for the optimal path. We restrict paths to only move one cell up or down at a time (or stay constant). For the boundary cases of $j=0,N_f$, we only search over $k \in \{0,1\}$ and $k \in \{-1,0\}$, respectively, to stay inside the grid (this is not shown in the pseudocode below).

### 2021-08-03 paper/ifo-introduction.tex
working on the response and minor changes to text

% CW 
To date, the network of gravitational-wave observatories has observed short-duration transient signals from the mergers of binary black holes, binary neutron stars, and neutron star-black hole mergers~\cite{GWTC-2:2020,NSBH:2021}.
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous-wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star in an LMXB can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous-wave searches. 
%Scorpius X-1 is an LMXB that is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~(e.g., Ref~\cite{ScoX1O2Viterbi:2019}).

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational-wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'', using light to capture sound, and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement. 
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~
\cite{SuvorovaEtAl:2017, ScoX1O2Viterbi:2019}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational-wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational-wave research groups around the world. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational-wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

This paper is laid out as follows. 
In Section~\ref{sec:ifo}, we detail the table-top interferometer design. 
In Section~\ref{sec:single_tone}, we demonstrate observing a single tone from a speaker. 
In Section~\ref{sec:viterbi_wandering}, we observe a wandering frequency signal and analyze it using a hidden Markov model technique (the Viterbi algorithm) from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capture and playback of complex audio, such as music and speech.
% removed as there's not space to really say what this means here: to varying degrees of success.
This demonstration of an optical microphone serves as a more general exhibition of signal processing with a range of examples that can be used in the undergraduate laboratory (described in the accompanying Supplementary Material). 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}. 
Further reading and resources are provided in the accompanying Supplementary Material and we present the software and scripts used to produce this work in Appendix~\ref{app:code}.

>>>>

% CW 
To date, the network of gravitational-wave observatories has observed short-duration transient signals from the mergers of binary black holes, binary neutron stars, and neutron star-black hole mergers~\cite{GWTC-2:2020,NSBH:2021}.
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals, which are yet to be detected. 
%Continuous waves may be emitted at a constant frequency or may wander slowly in frequency over time. 
Rotating neutron stars are prime candidates for continuous-wave emission, especially those in low mass X-ray binaries (LMXB), where the neutron star is in orbit with a low mass stellar companion.
%where a compact object such as a black hole or a neutron star is in orbit with a low mass stellar companion.
% we're talking about neutron stars in LMXB's, so we don't need to mention black holes in LMXB's.
The rotation frequency of the neutron star in an LMXB can wander over time due to variable accretion of matter (and hence angular momentum transfer) from the stellar companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is a prime LMXB target for continuous-wave searches. 
%Scorpius X-1 is an LMXB that is particularly bright in X-rays, making it a prime target for continuous wave searches. %from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~(e.g., Ref~\cite{ScoX1O2Viterbi:2019}).

% changed present to use, as this is not the first time a Michelson has been presented as a tpy GW detector - hannah
In this paper, we use a table-top Michelson interferometer as a toy gravitational-wave detector designed to detect sound instead of gravitational waves. 
We then extend its use to an ``optical microphone'', using light to capture sound, and present a range of example analysis techniques for educators to use. 
% re-added "as an undergraduate experiment to be clear we are not talking about school teaching - hannah
As an undergraduate lab experiment, the demonstration can be used to teach topics ranging from continuous-wave detection and analysis to electronics, signal processing, and speech enhancement.
% I don't think it is novel to use an optical device as a microphone, should we make this clearer? - james
%% I don't like the use of ``freedom to explore''. 
% removed Additionally - I don't think it is neccessary - hannah
It allows students in courses such as physics and electrical engineering to explore the response of an accessible, yet nontrivial, optomechanical system using a hierarchy of data analysis techniques of increasing complexity, including those used in the search for continuous waves in LIGO and Virgo data~
\cite{SuvorovaEtAl:2017, ScoX1O2Viterbi:2019}. 
% This demonstration can be adapted to an effective outreach tool to explain gravitational wave science to a non-specialist audience.
% Rephrased as this kind of sounds like we are an exhibition too [hannah] - This demonstration also has the potential to be used as an outreach tool alongside other exhibitions
%This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities including; hands-on demonstrations and exhibitions~\cite{TTExhibit:2020,CavagliaExhibit:2009,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, phone applications~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online} which have been developed by gravitational-wave research groups around the world. 
This demonstration also has the potential to be used as an outreach tool alongside a range of other public engagement demonstrations and activities developed by gravitational-wave research groups around the world. 
These tools allow scientists to cater to the increased public and media interest in this field and explain gravitational-wave science to non-specialist audiences.
%\han{look up Sarah Farmer's project}

This paper is laid out as follows. 
In Section~\ref{sec:ifo}, we detail the table-top interferometer design. 
In Section~\ref{sec:single_tone}, we demonstrate observing a single tone from a speaker. 
In Section~\ref{sec:viterbi_wandering}, we observe a wandering frequency signal and analyze it using a hidden Markov model technique (the Viterbi algorithm) from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capture and playback of complex audio, such as music and speech.
% removed as there's not space to really say what this means here: to varying degrees of success.
This demonstration of an optical microphone serves as a more general exhibition of signal processing with a range of examples that can be used in the undergraduate laboratory (described in the Supplementary Material). % deleted ``accompaning'', the material is already supplementary - james
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}. 
Further reading and resources are provided in the Supplementary Material and we present the software and scripts used to produce this work in Appendix~\ref{app:code}.

### 2021-08-03 paper/ifo-supplementary-material.tex

\title{Supplementary Material for \\Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone}

@@

@@

The methods used in Section~IV are inspired by hidden Markov model (HMM) searches for continuous gravitational waves. 
The HMM method is described in Refs~\cite{SuvorovaEtAl:2017,SuvorovaEtAl:2016,BayleyEtAlSOAP:2019,Viterbi:1967} and further information in Ref.~\cite{JKS:1998}. 
For more information on the Viterbi algorithm, the interested reader is referred to the excellent review paper by Rabiner~\cite{Rabiner:1989} and textbook by Quinn and Hannam~\cite{QuinnEtAl:2001} to learn more. 
Alternative pseudocode examples (to that described the appendix) can be found online at Ref.~\cite{viterbipseudocode:online}.

@@

\subsection{Environmental noise in gravitational-wave detectors}
In Section~V~B in the main article, we see evidence for environmental noise in our data. 
In gravitational-wave detectors reducing and mitigating environmental noise is an active area of research~\cite{EfflerEtAl:2015}. 
One such noise source is due to seismic ground motion from the Earth. 
In gravitational-wave observatories, the optical components are suspended from a series of pendulums to achieve passive isolation, and feedback control loops provide active isolation\cite{MatichardEtAl:2015, AdvancedVirgo:2015}. 
Such a system is beyond the scope of this article, however feedback control has been demonstrated in table-top Michelson interferometers designed for the undergraduate laboratory~\cite{LibbrechtBlack:2015}.
Other sources of noise include acoustic sound waves (for example from nearby vehicles, wind, and components such as electric fans) and magnetic noise (for example from high-voltage power lines, building heaters and lights, or any electrical power circuit)~\cite{EfflerEtAl:2015,CironeEtAl:2019}.
In some cases, environmental noise sources can be removed or mitigated on site~\cite{CovasEtAl:2018}. 
Where this is not possible, noise subtraction can be done after data collection~\cite{DriggersEtAl:2019, DavisEtAl:2019}. 
One example is the removal of spectral noise lines from the United States power grid operating at $60\,{\rm Hz}$~\cite{VajenteEtAl:2020}. %United States's?

@@

This work is implemented in Python Python 3~\cite{python} scripts and jupyter notebooks~\cite{jupyter,ipython} as well as in MATLAB~\cite{MATLAB}. 
We make use of Python packages: numpy~\cite{numpy}\footnote{\url{https://numpy.org/doc/stable/reference/routines.fft.html}}, scipy~\cite{scipy}, matplotlib~\cite{matplotlib}, tqdm~\cite{tqdm}, and logmmse~\cite{logmmse}.

@@

All analysis is done using the photodiode set-up as described in Section~V.A. in the main article. 
A photograph of the equipment is shown in Fig.~\ref{fig:photographs_of_optical_microphone_and_circuit} with the interferometer and circuit shown on the left and right, respectively. 
% schematic of breadboard and connections to pi
% https://www.circuit-diagram.org/editor/
% https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce
The photodiode circuit diagram is shown in Figure~\ref{fig:circuit_diagram}
The design was based off standard examples found online of photo-detectors~\cite{photodiodecircuits1:online,photodiodecircuits2:online}, connecting an analogy-to-digital converter ADC to a Pi~\cite{piadcguide:online}, and Sallen-Key filters~\cite{sallenkey1:online,sallenkey2:online}. Useful throughout was the Raspberry Pi documentation~\cite{pidocumentation:online} and the pinoutXYZ reference~\cite{pipinout:online}.

@@

\begin{figure}
	\begin{center}
    %% trim order is <left> <lower> <right> <upper>
    % height=0.5\textheight,trim={1cm 4.5cm 0.4cm 3.5cm},clip
	\includegraphics[width=.5\textwidth]
% trim={1cm 3cm 1cm 1cm}]
{figures/combined_results_full_melatos_labelled.pdf}
%notch_and_wiener_superplot_v2.pdf}
	\end{center}
	\caption{\label{fig:op_mic_results}
	%\jam{Fig.2 should be combined with Fig.6 using combined\_results\_full\_melatos\_labelled.pdf}
	Timeseries (left column) and frequency spectrum (right column) results for the background subtraction, notch, and Wiener filters.
    The top two rows are the input signal and raw output (identical to the top two rows in Fig.~7 in the main article).  
    The third row shows a background noise recording from the optical microphone when no audio signal is played. 
    The fourth row shows the result of subtracting the background noise spectrum in the third row from the recording in the second row. 
	The fifth and sixth rows show the results of applying the cascaded-notch filter and Wiener filter respectively.
    The seventh shows the results of applying both the cascaded-notch and Wiener filter (identical to the third row in Fig.~7 in the main article).
    The eighth row shows the results of applying the logMMSE estimator (identical to the fourth row in Fig.~7 in the main article). 
	}
\end{figure}

Results for a selection of filters are collated in Fig.~\ref{fig:op_mic_results}, which we refer to throughout this Supplementary Material. 
The left and right columns show timeseries and Fourier spectrum results, respectively. 
The first and second rows in Fig.~\ref{fig:op_mic_results} show the original source signal and the raw optical microphone recording, respectively. 
The following sections describe a sequence of analysis techniques and a selection of their results is presented in Fig.~\ref{fig:op_mic_results}.

@@

A Wiener filter is an advanced statistical technique that estimates the injected signal given prior information about the injected spectrum and the reference spectrum of the background noise. 
The observed noisy speech signal sequence is given as $\mathbf{x}=(x(0),\dots, x(N-1))$, where $N$ is the length of the data sequence and $\mathbf{x}$ is the sum of the original injected signal $\mathbf{s}=(s(0),\dots,s(N-1))$, and the noise sequence $\mathbf{w}=(w(0),\dots,w(N-1))$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}\,.
\end{equation}
Given $\textbf{x}$, our goal is to make an estimate $\hat{\textbf{s}}$ of the original signal $\textbf{s}$ such that we minimise the Bayesian mean-square error (BMSE) between the two, defined as 
\begin{equation}
\label{eq:BMSE}
\text{BMSE}(\hat{\textbf{s}})=E[(\textbf{s}-\hat{\textbf{s}})^2]\,.
\end{equation}
If we assume that: i) $\textbf{x}$ is ``wide sense stationary''~\footnote{ A random process $\{x(t)\}$ is wide sense stationary if, for all $t_1,t_2 \in \mathbb{R}$, (1) its mean is time invariant, i.e., $\mu_x(t_1)=\mu_x(t_2)=\text{constant}$; and (2) the autocorrelation depends only on the time difference, i.e., $R_x(t_1,t_2)=R_x(\tau),\tau=t_1-t_2$.} with zero mean; ii) the signal $\textbf{s}$ has a mean of zero; and iii) the noise $\textbf{w}$ is uncorrelated with the signal $\textbf{s}$, we can further express $\hat{\textbf{s}}$ to be a linear combination of present and past observed data
\begin{equation}
\hat{{s}}[n]=\sum_{k=0}^{n}h[k]x[n-k]\,,
\end{equation}
where $\textbf{h}=(h(0),\dots h(n))$ represents the coefficients of an $n$th order Wiener filter.
The famous Wiener-Hopf equation~\citep{noble1959methods} allows us to determine $\textbf{h}$, as\begin{equation}
\label{eqn:wiener-hopf}
\begin{bmatrix}  
r_{xx}[0]&r_{xx}[1]&\dots& r_{xx}[n]\\
r_{xx}[1]&r_{xx}[0]&\dots &r_{xx}[n-1]\\
\vdots&\vdots&\ddots&\vdots\\
r_{xx}[n]&r_{xx}[n-1]&\dots &r_{xx}[0]
\end{bmatrix}
\begin{bmatrix}
h[0]\\
h[1]\\
\vdots\\
h[n]
\end{bmatrix}=
\begin{bmatrix}
r_{ss}[0]\\
r_{ss}[1]\\
\vdots\\
r_{ss}[n]
\end{bmatrix}\,,
\end{equation}
where $r_{xx}$ and $r_{ss}$ are the auto-correlation functions of $\mathbf{x}$ and $\mathbf{s}$ between timestep $i$ and $i+n$, 
%\begin{equation*}
%\left\{ 
\begin{eqnarray} 
r_{xx}[n] &~=~& E[x(i)~x(i+n)] \,,\\
r_{ss}[n] &~=~& E[s(i)~s(i+n)] \,.
%r_{xx}[n] &=& E[x(i)\, &x(i+n)] \\ 
%r_{ss}[n] &=& E[s(i)\, &s(i+n)] 
\end{eqnarray} 
%\right\}  
%\end{equation*}

If we let the Wiener filter be non-causal (i.e. we estimate the current signal based on both past \emph{and future} observations), then we can represent Eqn.~\ref{eqn:wiener-hopf} in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}\,,
\end{equation}
where $P_{xx}(f), P_{ss}(f), P_{ww}(f)$ are the spectra of the observed noisy data, the injected signal, and the background noise, respectively. Intuitively, from Eqn.~\ref{eqn:wiener}, we can see that the non-causal Wiener filter amplifies the input signal where the signal to noise ratio (SNR) is high and attenuates the signal where the SNR is low. The causal Wiener filter (which only makes estimates based on past observations) is similar. A more detailed analysis of both kinds of Wiener filters can be found in Ref.~\citep{10.5555/151045}.

>>>>

\title{Supplementary Material for \texorpdfstring{\\}{}Continuous gravitational waves in the lab: recovering audio signals with a table-top~optical microphone}

@@

\date{\today}

@@

The methods used in Section~IV are inspired by hidden Markov model (HMM) searches for continuous gravitational waves. 
The HMM method is described in Refs~\cite{SuvorovaEtAl:2017,SuvorovaEtAl:2016,BayleyEtAlSOAP:2019,Viterbi:1967} and further information in Ref.~\cite{JKS:1998}. 
For more information on the Viterbi algorithm, the interested reader is referred to the excellent review paper by Rabiner~\cite{Rabiner:1989} and textbook by Quinn and Hannam~\cite{QuinnEtAl:2001} to learn more. 
Alternative pseudocode examples (to that described in Appendix~C of the main paper) can be found online at Ref.~\cite{viterbipseudocode:online}.

@@

\subsection{Environmental noise in gravitational-wave detectors}
In Section~V~B in the main article, we see evidence for environmental noise in our data. 
In gravitational-wave detectors reducing and mitigating environmental noise is an active area of research~\cite{EfflerEtAl:2015}. 
One such noise source is due to seismic ground motion from the Earth. 
In gravitational-wave observatories, the optical components are suspended from a series of pendulums to achieve passive isolation, and feedback control loops provide active isolation\cite{MatichardEtAl:2015, AdvancedVirgo:2015}. 
Such a system is beyond the scope of this article, however, feedback control has been demonstrated in table-top Michelson interferometers designed for the undergraduate laboratory~\cite{LibbrechtBlack:2015}.
Other sources of noise include acoustic sound waves (for example from nearby vehicles, wind, and components such as electric fans) and magnetic noise (for example from high-voltage power lines, building heaters and lights, or any electrical power circuit)~\cite{EfflerEtAl:2015,CironeEtAl:2019}.
In some cases, environmental noise sources can be removed or mitigated on site~\cite{CovasEtAl:2018}. 
Where this is not possible, noise subtraction can be done after data collection~\cite{DriggersEtAl:2019, DavisEtAl:2019}. 
One example is the removal of spectral noise lines from the United States power grid operating at $60\,{\rm Hz}$~\cite{VajenteEtAl:2020}. %United States's?

@@

This work is implemented in Python 3~\cite{python} scripts and jupyter notebooks~\cite{jupyter,ipython} as well as in MATLAB~\cite{MATLAB}. 
We make use of Python packages: numpy~\cite{numpy}\footnote{\url{https://numpy.org/doc/stable/reference/routines.fft.html}}, scipy~\cite{scipy}, matplotlib~\cite{matplotlib}, tqdm~\cite{tqdm}, and logmmse~\cite{logmmse}.

@@

All analysis is done using the photodiode set-up as described in Section~V.A. in the main article. 
A photograph of the equipment is shown in Fig.~\ref{fig:photographs_of_optical_microphone_and_circuit} with the interferometer and circuit shown on the left and right, respectively. 
% schematic of breadboard and connections to pi
% https://www.circuit-diagram.org/editor/
% https://crcit.net/c/e397dcc2166943d69155f9dac1e27bce
The photodiode circuit diagram is shown in Figure~\ref{fig:circuit_diagram}
The design was based off standard examples found online of photo-detectors~\cite{photodiodecircuits1:online,photodiodecircuits2:online}, connecting an analogy-to-digital converter ADC to a Raspberry~Pi~\cite{piadcguide:online}, and Sallen-Key filters~\cite{sallenkey1:online,sallenkey2:online}. Useful throughout was the Raspberry Pi documentation~\cite{pidocumentation:online} and the pinoutXYZ reference~\cite{pipinout:online}.

@@

\begin{figure}
	\begin{center}
    %% trim order is <left> <lower> <right> <upper>
    % height=0.5\textheight,trim={1cm 4.5cm 0.4cm 3.5cm},clip
	\includegraphics[width=.48\textwidth]
% trim={1cm 3cm 1cm 1cm}]
{figures/combined_results_full_melatos_labelled.pdf}
%notch_and_wiener_superplot_v2.pdf}
	\end{center}
	\caption{\label{fig:op_mic_results}
	%\jam{Fig.2 should be combined with Fig.6 using combined\_results\_full\_melatos\_labelled.pdf}
	Timeseries (left column) and frequency spectrum (right column) results for the background subtraction, notch, and Wiener filters.
    The top two rows are the input signal and raw output (identical to the top two rows in Fig.~7 in the main article).  
    The third row shows a background noise recording from the optical microphone when no audio signal is played. 
    The fourth row shows the result of subtracting the background noise spectrum in the third row from the recording in the second row. 
	The fifth and sixth rows show the results of applying the cascaded-notch filter and Wiener filter respectively.
    The seventh shows the results of applying both the cascaded-notch and Wiener filter (identical to the third row in Fig.~7 in the main article).
    The eighth row shows the results of applying the logMMSE estimator (identical to the fourth row in Fig.~7 in the main article). 
	}
\end{figure}

Results for a selection of filters are collated in Fig.~\ref{fig:op_mic_results}, which we refer to throughout this Supplementary Material. 
The left and right columns show timeseries and Fourier spectrum results, respectively. 
The first and second rows in Fig.~\ref{fig:op_mic_results} show the original, source signal and the raw optical microphone recording, respectively. 
The following sections describe a sequence of analysis techniques and a selection of their results is presented in Fig.~\ref{fig:op_mic_results}.

@@

A Wiener filter is an advanced statistical technique that estimates the injected signal given prior information about the injected spectrum and the reference spectrum of the background noise. 
The observed noisy speech signal sequence is given as $\mathbf{x}=(x(0),\dots, x(N-1))$, where $N$ is the length of the data sequence and $\mathbf{x}$ is the sum of the original injected signal $\mathbf{s}=(s(0),\dots,s(N-1))$, and the noise sequence $\mathbf{w}=(w(0),\dots,w(N-1))$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}\,.
\end{equation}
Given $\textbf{x}$, our goal is to make an estimate $\hat{\textbf{s}}$ of the original signal $\textbf{s}$ such that we minimise the Bayesian mean-square error (BMSE) between the two, defined as 
\begin{equation}
\label{eq:BMSE}
\text{BMSE}(\hat{\textbf{s}})=E[(\textbf{s}-\hat{\textbf{s}})^2]\,.
\end{equation}
If we assume that: (i) $\textbf{x}$ is ``wide sense stationary''~\footnote{ A random process $\{x(t)\}$ is wide sense stationary if, for all $t_1,t_2 \in \mathbb{R}$, (1) its mean is time invariant, i.e., $\mu_x(t_1)=\mu_x(t_2)=\text{constant}$; and (2) the autocorrelation depends only on the time difference, i.e., $R_x(t_1,t_2)=R_x(\tau),\tau=t_1-t_2$.} with zero mean; (ii) the signal $\textbf{s}$ has a mean of zero; and (iii) the noise $\textbf{w}$ is uncorrelated with the signal $\textbf{s}$, we can further express $\hat{\textbf{s}}$ to be a linear combination of present and past observed data
\begin{equation}
\hat{{s}}[n]=\sum_{k=0}^{n}h[k]x[n-k]\,,
\end{equation}
where $\textbf{h}=(h(0),\dots h(n))$ represents the coefficients of an $n$th order Wiener filter.
The famous Wiener-Hopf equation~\citep{noble1959methods} allows us to determine $\textbf{h}$, as\begin{equation}
\label{eqn:wiener-hopf}
\begin{bmatrix}  
r_{xx}[0]&r_{xx}[1]&\dots& r_{xx}[n]\\
r_{xx}[1]&r_{xx}[0]&\dots &r_{xx}[n-1]\\
\vdots&\vdots&\ddots&\vdots\\
r_{xx}[n]&r_{xx}[n-1]&\dots &r_{xx}[0]
\end{bmatrix}
\begin{bmatrix}
h[0]\\
h[1]\\
\vdots\\
h[n]
\end{bmatrix}=
\begin{bmatrix}
r_{ss}[0]\\
r_{ss}[1]\\
\vdots\\
r_{ss}[n]
\end{bmatrix}\,,
\end{equation}
where $r_{xx}$ and $r_{ss}$ are the auto-correlation functions of $\mathbf{x}$ and $\mathbf{s}$ between timestep $i$ and $i+n$, 
%\begin{equation*}
%\left\{ 
\begin{eqnarray} 
r_{xx}[n] &~=~& E[x(i)~x(i+n)] \,,\\
r_{ss}[n] &~=~& E[s(i)~s(i+n)] \,.
%r_{xx}[n] &=& E[x(i)\, &x(i+n)] \\ 
%r_{ss}[n] &=& E[s(i)\, &s(i+n)] 
\end{eqnarray} 
%\right\}  
%\end{equation*}

If we let the Wiener filter be non-causal (i.e. we estimate the current signal based on both past \emph{and future} observations), then we can represent Eqn.~\ref{eqn:wiener-hopf} in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}\,,
\end{equation}
where $P_{xx}(f), P_{ss}(f), P_{ww}(f)$ are the spectra of the observed noisy data, the injected signal, and the background noise, respectively. Intuitively, from Eqn.~\ref{eqn:wiener}, we can see that the non-causal Wiener filter amplifies the input signal where the signal-to-noise ratio (SNR) is high and attenuates the signal where the SNR is low. The causal Wiener filter (which only makes estimates based on past observations) is similar. A more detailed analysis of both kinds of Wiener filters can be found in Ref.~\citep{10.5555/151045}.
